Towards a Reinforcement Learning approach for Aggregate Computing

### 2021-11-29 paper-2021-woa-ac-ml.tex
final Musolesi pass

\section{Introduction}
Recent trends like pervasive~\cite{DBLP:journals/wc/Satyanarayanan01} and autonomic~\cite{DBLP:journals/computer/KephartC03} computing 
 promote a vision of artificially designed Collective Adaptive Systems (CASs)~\cite{DBLP:conf/huc/Ferscha15},
 namely collections of agents coordinating and adapting %as wholes 
 to pursue joint goals in changing environments.
% due to the increasing interconnected computational devices placed in ever-changing environments.
 Examples of CASs include robot swarms, smart cities, and digitally-augmented crowds.
%
%CASs are characterised by 
%\begin{itemize}
%\item a global order that emerges from local node interactions, 
%\item  decentralised control, and 
%\item adaptive behaviours towards environmental changes to pursuit a collective goal.
%\end{itemize}
%
CAS are notoriously hard to design and control 
 %due to as part of complex systems
 as globally desired behaviour
 has to emerge from local behaviours and interactions. 
%
For this reason, over the years, various approaches have tried to handle this complexity---for instance, by taking inspiration from natural systems like insect and animal societies (e.g., ant colonies, bee hives, flocks of birds etc.). 
%
%In this case, a sort of hive mind arises 
% %-- also referred to as swarm intelligence --  
% born from local interactions between agents. 
%
Specifically, such approaches tend to exploit 
 \emph{self-organisation}, 
 i.e., processes by which global order emerges 
 from continuous interactions among simple entities.

Initially, researchers had tried to achieve these properties in computing systems by miming natural collective behaviours ``from the bottom up''~\cite{DBLP:journals/connection/Webb02}, 
 namely trying to find the right single-node-behaviour (local rules)
 to pursue the intended global behaviour. %, e.g. flocking, foraging, etc.
%
However, such local rules cannot easily be defined, since emergence, and its effects, are (almost by definition) hard to be controlled or predicted in complex systems.
%
A different approach could attempt to express collective behaviours in a more top-down fashion, in terms of the desired emergents.
%, abstracting over non-functional aspects such as the network topology, environment uncertainties and system size.
In this direction, Aggregate Computing (AC)~\cite{DBLP:journals/computer/BealPV15} is proposed:
 it is a paradigm by which
 developers can express collective self-organising behaviours from a global perspective. % through a functional program specification.
%
Indeed, an aggregate program 
 (i.e. a program written with AC) 
 consists of the manipulation of \textit{computational fields}, i.e., distributed
 data structures where each node is associated with the result of a computation performed by it,
 and neighbourhood observation.
%
%These operations are defined in \textit{field-calculus}~\cite{DBLP:conf/coordination/AudritoBDV18} --- the root of Aggregate Computing. 
% This calculus describes a set of minimal constructs to express any spatiotemporal computation.

AC captures common abstractions for collective behaviour specification in so-called ``building-blocks''. 
 %These abstractions are encapsulated in so-called ``building blocks''.
 %through which various applications have been built (e.g., for swarm robotics and crowds engineering~\cite{DBLP:journals/eaai/CasadeiVAPD21}).
However, building block synthesis concerns different stuff not only related to the behaviour itself
 but also to the ``ensemble" specifications. 
%
For instance, it is quite difficult to craft building blocks
 that easily adapt w.r.t. the node movements rapidity.
%
Ideally, we would like to continue building our application using these building blocks 
 -- so the high-level specification remains the same -- 
 but using some under-the-hood mechanisms that adapt the behaviour according to different environmental changes. 
%
For this reason, we claim that Machine Learning can help us in handling this "unseen" situation, leaving only the burden
 to developers of defining the right collective specifications.
%This lead to an ``hybrid" approach where a part of behaviour is still expressed using common 
% AC abstractions and another part is distilled through learning~\cite{research}.
In particular, in this paper,
 we investigate how AC can be extended with Reinforcement Learning (RL) through a case study,
 where we outline the performance improvement of the ``hop count" program ---
 i.e. an aggregate program that produces a field that stores the distance from a source zone. 
 Furthermore, we show that the learnt policy scales up to different system sizes.

@@

Collective Adaptive Systems are characterized by a very high number of nodes without a central coordinating authority that coordinate nodes. 
 The overall behaviour is a consequence of local interactions.%The systems expose self-* properties.

Engineering this kind of system is a thriving area, and so various approaches have been proposed.
 First attempts leverage \textit{local-to-global} techniques, where the collective behaviour emerges from local behaviour description, 
 usually inspired by nature~\cite{DBLP:journals/swarm/BrambillaFBD13}. 
%
Instead, a natural way to describe this kind of problems should be through a \textit{collective} view of the system.
 This is the idea used by the \textit{global-to-local} techniques~\cite{DBLP:journals/jlap/ViroliBDACP19,DBLP:journals/scp/AlrahmanNL20, DBLP:conf/cbse/BuresGHKKP13}, 
 where the behaviours are designed \textit{top-down}, and then the local behaviour is generated automatically from this ``system-wide'' specification. 
% 
Also ``meet-in-the-middle''~\cite{DBLP:journals/computer/PinciroliB16} approaches exist, where
 a part of the system is still expressed bottom-up (device-centric) whereas a part is expressed top-down (swarm-centric)

In this work, we focus on Aggregate Computing~\cite{DBLP:journals/computer/BealPV15}, a global-to-local approach where
 the self-organising collective behaviour is expressed in terms of manipulation of \textit{computational fields}.
The latter could be seen as the extension of electromagnetic fields to Computer Science since it consists of a 
 map from any devices to their computational value (i.e. the result of a computation).
The computational field manipulation is ruled by \textit{field calculus}, which define the constructs needed 
 to express spatio-temporal computations~\cite{DBLP:journals/jlap/ViroliBDACP19}.
%
The novelty of AC consists in its composable (functional-like) system-size independent program definition: 
 using basic constructs we identify the main patterns (\textit{building-blocks}) that could be reusable at the application level.
By opportunistically crafting these building-blocks it is possible to verify prominent properties such as eventual consistency %~\cite{DBLP:journals/taas/BealVPD17} 
 (i.e. the ensemble behaviour is independent of the underlying network details)
 and self-stabilisation %~\cite{DBLP:journals/tomacs/ViroliABDP18} 
 (i.e. the ability of a system to recover from arbitrary changes)~\cite{DBLP:journals/jlap/ViroliBDACP19}.

@@

\subsection{Field Calculus operators}
The main constructs that capture the essential aspects for programming self-organising systems with field calculus are:
\begin{itemize}
  \item \textit{Stateful field evolution} expression \mintinline[escapeinside=||]{scala}{|{\color{teal}rep}|(|$e_1$|) {|($x$) => $e_2$|}} describes a field evolving in time. 
  $e_1$ is the initial field value and the function $(x) => e_2$ define how the field changes at each execution.
  \item \textit{Neighbour interaction} expression \mintinline[escapeinside=||]{scala}{|\color{teal}nbr|{|$e$|}} a view
  of the field values in the surroundings of each device where neighbours are
  mapped to their evaluations of $e$. A series of \mintinline{scala}{*hood} operators leverage this map to produce relevant result. 
  For instance, \mintinline{scala}{minHood} return the minimum value 
  within the neighbourhood.
  \item \textit{Domain partitioning} expression \mintinline[escapeinside=||]{scala}{|\color{teal}if|(|$e_0$|){|$e_1$|}{|$e_2$|}} splits the computational
  field into two non-communicating zones hosting isolated subcomputations:
  $e_1$ where $e_0$ is true, and $e_2$ where $e_0$ is false.
\end{itemize}
%
A full coverage of the field calculus is beyond the scope of this paper. A more comprehensive presentation is given in~\cite{DBLP:journals/jlap/ViroliBDACP19}.

@@

In this case, when Reinforcement Learning is applied to several agents, we speak of Multi-Agent Reinforcement Learning (MARL)~\cite{DBLP:journals/aamas/Hernandez-LealK19}.
 AC has a typical cooperative and homogeneous learning setting: 
 each agent collaborates to achieve a common goal guided by the same reward function.
%
In the cooperative and homogeneous systems, several works have tried to apply the algorithms used in the single-agent context to the multi-agent one --- 
 the so-called Independent MARL~\cite{DBLP:journals/tsmc/BusoniuBS08}. 
%
However, several issues emerge, such as non-stationarity, multi-agent credit assignment, scalability and partial observability of the system. 
 Therefore, several heuristics have tried to solve these problems.
%
Among many, Hysteric Q-Learning~\cite{hysteretic-q} uses an optimistic approach in which more weight is given to positive actions than negative ones.
%
Another way to handle multi-agent learning is to use so-called \textit{Centralised Training and Decentralised Execution} (CTDE) algorithms. 
 In this case, a central agent collects the experiences of the various agents and then distils a policy that each agent can execute locally in a typical actor-critic setting. 
%
During the learning phase, the problems of partial observability and non-stationarity are reduced, since the centralised learner has a global view. 
 However, the problem of scalability is typically not addressed.

@@

\section{Aggregate Computing with Reinforcement}\label{aggregate-and-rl}
\begin{figure}
  \centering
  \includegraphics[width=\linewidth]{img/algorithm-learning.pdf}
  \caption{This figure shows the main steps of RL in AC.
  Each agent starts with his copy of a Q table. 
  Then, during the simulation, they refine them Q tables changing Q values according to their experience (different colours mean different values). 
  In the end, a new $Q_{max}$ table will be created aggregating each agent table. 
  Finally, a new episode starts using this $Q_{max}$ as a starting Q table.
  }
  \label{fig:aggregate-q-learning}
\end{figure}
%As first experiments, we leverage RL to improve building blocks. 
% 
In this section, we briefly describe how RL can be injected into the AC computation loop,
 and what RL we used with AC.

From a local view, AC program execution can be seen as:
$$
f : \textit{Context} \rightarrow \textit{Output}
$$
Where $\textit{Context}$ contains:
\begin{itemize}
  \item $\sigma(name) \rightarrow \textit{data}$: a function from sensor name to the perceived value;
  \item $\phi(\textit{ID}) \rightarrow \textit{Export}$: a function from id to neighbourhoods export data.
\end{itemize}
The computational field $\theta$ contains all output produced by the system.

Speaking of RL settings, we need to define the state function and the reward function.
%
The execution needs to be distributed, so the state will be a function from the local node context:
$$
K := \Omega(\textit{Context})
$$

Reward function instead, might be depending on the learning technique, time, and problem. 
 Indeed, if we perform learning offline, it could be defined from the overall \textit{simulated} computational field:
$$
\textit{R}_{global} := \psi({\theta})
$$
If we use an online mode, it needs to be defined as a function to the local node context:
$$
\textit{R}_{local} := \psi(\textit{Context})
$$
Finally, actions influence the current aggregate program output, changing the program evaluation as:
$$
f_{rl} : \textit{Context} \rightarrow \lambda(\textit{Output}, \textit{Action})
$$
As remarked before, the policy $\pi$ need to be distributed, so each agent can choose the action that will be performed:
$$
\textit{Action} \leftarrow \pi(K)
$$

As a first step towards a complete integration between AC and RL, we have decided to set up an Independent RL learning technique, 
 by which each agent try to improve their own current local policy (shown in \Cref{fig:aggregate-q-learning}). 
% 
We made this choice inspired by other work in CSAS~\cite{csas-and-marl} and swarm systems~\cite{nguyen2018swarm}.

Our approach is partially inspired by Swarm Q-Learning~\cite{nguyen2018swarm} and leverage Hysteretic Q-Learning~\cite{hysteretic-q} to
 handle the non-stationarity in the environment.
%
In particular, each agent updates a local Q table ($Q_i(s, a)$, where $i$ is the agent id) following the standard hysteretic update:
$$
\delta \leftarrow r + \gamma * max_a Q_i(s', a) - Q_i(s_t, a_t)
$$
$$
Q(s_t, a_t) =  \begin{cases} 
  Q(s_t, a_t) + \alpha * \delta & \mbox{if } delta >= 0 \\ 
  Q(s_t, a_t) + \beta * \delta & \mbox{otherwise }
\end{cases}
$$

At the end of each episode, each agent has collected its local experience, which can vary in different zones. 
 So, we combine the N (the number of nodes) Q table selecting the maximum value for each state-action pair since we assume an optimistic homogeneous behaviour:
$$
Q_{max}(s, a) = max_i(Q_i(s, a))
$$
Then, in the next episode, each agent starts with the $Q_{max}$ and try to refine it.
\section{Evaluation}\label{evaluation}
\begin{figure*}[h]
  \centering
  \begin{subfigure}[b]{0.32\textwidth}
      \centering
      \includegraphics[width=\textwidth]{img/hop-count-1.png}
      \caption{Stable situation: Two source nodes exists}
  \end{subfigure}
  \hfill
  \begin{subfigure}[b]{0.32\textwidth}
      \centering
      \includegraphics[width=\textwidth]{img/hop-count-2.png}
      \caption{Unstable situation: One source disappears}
  \end{subfigure}
  \hfill
  \begin{subfigure}[b]{0.32\textwidth}
      \centering
      \includegraphics[width=\textwidth]{img/hop-count-3.png}
      \caption{Stable state reached}
  \end{subfigure}
  \caption{Graphical result of hop count evaluation. The node with a red cross is the target zone. The label 
  associated with each node is the result of the aggregate program evaluated in that zone.}
  \label{fig:hop-count}
\end{figure*}
In this section, we outline the results of the application of RL for one simple
 aggregate program: the hop count. In this case, the computational field that will eventually be produced consists of the distance in terms of the minimum number of nodes needed to reach source nodes (\Cref{fig:hop-count}).

@@

Where \textbf{\texttt{k}} and \textbf{\texttt{w}} are two parameters of this algorithm.
%
The policy is learnt locally, during simulation, using our version of Q-Learning algorithm. 
 The reward function consists in: 
\[
\systeme*{0 : & \mbox{\textit{hop count output is right}}, -1 : & \mbox{\textit{otherwise}}}
\]
The idea is to maintain the nodes in a stable situation or to
 fasten the time needed to reach a stable one.
%
\Cref{fig:aggregate-learning-loop} shows the learning loop in details.
\begin{figure}
  \centering
  \includegraphics[width=\linewidth]{img/aggregate-program-evaluation.pdf}
  \caption{Each agent, starting from their context, 
  periodically performs the standard hop count. 
  Using the output produced and the history of state and output, 
  each agent computes its current state. 
  Using it, they evaluate the action to speed up the hop count convergence following a policy. 
  This new refined output is then used to evaluate the reward and to improve the Q table (dotted line path)}
  \label{fig:aggregate-learning-loop}
\end{figure}

\subsection{Simulation Setup}
We verify our learning algorithm through simulation\footnote{repository at \url{https://github.com/cric96/alpaca-2022-ac-rl}} leveraging Alchemist and ScaFi.
 To check if the aggregate eventually speeds up the hop count convergence time, we create a situation where the rising problem emerges (\Cref{fig:hop-count}):
 firstly, two source nodes exist in the network, ($A$ and $B$).
 At the time $t$, $B$ disappears. So the nodes near $B$ start a slow process
 to eventually reach the distance towards $A$.
We use 80 ($N$) nodes displaced into an irregular grid with a width of 100 meters and a height of 20 meters. 
 Each node has a neighbourhoods range equal to 8 meters --- on average, each node has 7 neighbours. 
 The time $t$ is fixed to 20.
 Each simulation episode lasts 60-time unit. Finally, we set $k = 3$ and $w = 2$.
% 
We train the aggregate for 500 episodes.
 The Q-learning parameters are:
$$ 
\alpha = 0.5,
\beta = 0.05,
\gamma = 0.9
$$
Moreover, we use a $\epsilon$ decay policy for the $\epsilon$-greedy policy as:
\begin{equation*}
  \begin{array}{l}
    \epsilon_t = \epsilon_{start} * e^{-t/\theta}\\ 
    \epsilon_{start }= 0.99, \theta=4000
  \end{array}
\end{equation*}
In doing so, the agent will explore the state-action space as much as possible.

@@

Then, we test our learnt algorithm in different environments (\Cref{fig:simulation} b,c,d). To do that, we share the learnt Q-table to the 
 entire aggregate and we test the program changing the system size (80, 360 and 760 nodes).
%
For each population size, we run 50 episodes and then we average the error at each time step. 
 As a result, the aggregate speed-up convergence in each new situation.

@@

\section{Discussion and Concluding Remarks}\label{conclusion}
In this article, we have discussed the combination of AC with Machine Learning techniques in order to improve adaptivity
 and effectiveness of the AC framework.
%
In particular, we explore the combination of AC with Hysteretic Q-Learning.
%
RL application in this kind of setting is quite unexplored %-- even if some works exists~\cite{DBLP:conf/icml/YangLLZZW18} -- 
 area for several reasons, such as the high system dynamic, non-stationarity,  partial environment observability, and non-bounded system size.
In a case study, we empirically show the effectiveness of RL, by which we speed up the convergence time of
 aggregate program taking into consideration -- the hop count.
Furthermore, we have shown that the learnt program can easily scale up to system size.

To the best of our knowledge, this is the first effort towards a ``hybrid'' collective programming approach, where the high-level program specification is 
 done by a designer and the low-level matters are left to machine learning algorithms.
%
As such, this work shed light on several issues deserving future investigation.

Aggregate systems are partially observable from the perspective of each node. 
 Local observations may not be statistically significant in summarising the state of the entire aggregate. 
 In the future, we could address this problem through recurrent neural networks and a CTDE learning methodology.

The learning, in this case, was applied at the building block level. 
 Another possible direction is to use it at the framework level, e.g., to select which messages to send to the neighbourhood or not. 
 This has already been addressed in the literature under the name of \textit{learn to communicate}~\cite{DBLP:journals/corr/abs-1908-03963}. 
A novel way to handle this problem is to use Graph Neural Networks~\cite{DBLP:journals/corr/abs-1812-08434} to learn the representation of a distributed system and to select messages based on that representation. 
 Thus, we should consider this methodology when we will extend RL at the AC framework level.

In our experiment, we have applied a variation of the independent learning approach. 
 Agents first learn the best policy locally, and then at the end of the episode, a learner merges the Q-tables. 
 However, this does not reduce the problems of scalability and non-stationary, since a central agent must still collect all the Q-tables of the aggregate. 
 In particular, to handle the scalability issues, we will have to consider decentralised approaches based only on neighbour data (or local leaders).

>>>>

\section{Introduction}
Recent trends like pervasive~\cite{DBLP:journals/wc/Satyanarayanan01} and autonomic~\cite{DBLP:journals/computer/KephartC03} computing 
 promote a vision of artificially designed Collective Adaptive Systems (CASs)~\cite{DBLP:conf/huc/Ferscha15},
 namely collections of agents coordinating and adapting %as wholes 
 to pursue joint goals in changing environments.
% due to the increasing interconnected computational devices placed in ever-changing environments.
 Examples of CASs include robot swarms, smart cities, and digitally-augmented crowds.
%
%CASs are characterised by 
%\begin{itemize}
%\item a global order that emerges from local node interactions, 
%\item  decentralised control, and 
%\item adaptive behaviours towards environmental changes to pursuit a collective goal.
%\end{itemize}
%
CAS are notoriously hard to design and control 
 %due to as part of complex systems
 as globally desired behaviour
 has to emerge from local behaviours and interactions. 
%
For this reason, over the years, various approaches have addressed the problem of dealing with this complexity---for instance, by taking inspiration from natural systems like insect and animal societies (e.g., ant colonies, bee hives, flocks of birds etc.). 
%
%In this case, a sort of hive mind arises 
% %-- also referred to as swarm intelligence --  
% born from local interactions between agents. 
%
Specifically, such approaches tend to exploit 
 \emph{self-organisation}, 
 i.e., processes by which global order emerges 
 from continuous interactions among simple entities.

Initially, researchers had tried to build systems that show these properties in computing systems by mimickng natural collective behaviours ``from the bottom up''~\cite{DBLP:journals/connection/Webb02}, 
 namely trying to find the right single-node-behaviour (local rules)
 to achieve the intended global behaviour. %, e.g. flocking, foraging, etc.
%
However, such local rules cannot easily be defined, since emergence, and its effects, are (almost by definition) hard to be controlled or predicted in complex systems.
%
A different approach could attempt to express collective behaviours in a more top-down fashion, in terms of the desired emergents.
%, abstracting over non-functional aspects such as the network topology, environment uncertainties and system size.
In this direction, Aggregate Computing (AC)~\cite{DBLP:journals/computer/BealPV15} is proposed:
 it is a paradigm by which
 developers can express collective self-organising behaviours from a global perspective. % through a functional program specification.
%
Indeed, an aggregate program 
 (i.e. a program written with AC) 
 consists of the manipulation of \textit{computational fields}, i.e., distributed
 data structures where each node is associated with the result of a computation performed by it,
 and neighbourhood observation.
%
%These operations are defined in \textit{field-calculus}~\cite{DBLP:conf/coordination/AudritoBDV18} --- the root of Aggregate Computing. 
% This calculus describes a set of minimal constructs to express any spatiotemporal computation.

AC captures common abstractions for collective behaviour specification in so-called ``building-blocks''. 
 %These abstractions are encapsulated in so-called ``building blocks''.
 %through which various applications have been built (e.g., for swarm robotics and crowds engineering~\cite{DBLP:journals/eaai/CasadeiVAPD21}).
However, building block synthesis concerns different elements not only related to the behaviour itself
 but also to the ``ensemble" specifications. 
%
For instance, it is quite difficult to craft building blocks
 that easily adapt to the non stationarity due to node movements.
%
Ideally, we would like to continue developing our applications using these building blocks 
 -- so the high-level specification remains the same -- 
 but using some under-the-hood mechanisms that adapt the behaviour according to different environmental changes. 
%
For this reason, we claim that Machine Learning can help us in handling this "unseen" situation, leaving only the burden of defining the right collective specifications to developers.
%This lead to an ``hybrid" approach where a part of behaviour is still expressed using common 
% AC abstractions and another part is distilled through learning~\cite{research}.
In particular, in this paper,
 we investigate how AC can be extended with Reinforcement Learning (RL) through a case study,
 where we outline the performance improvement of the ``hop count" program ---
 i.e. an aggregate program that produces a field that stores the distance from a source zone. 
 Furthermore, we show that the learnt policy scales up to different system sizes.

@@

Collective Adaptive Systems are characterized by a very high number of nodes without a central authority that coordinate nodes. 
 The overall behaviour is a consequence of local interactions.%The systems expose self-* properties.

Engineering this kind of system is a thriving area, and so various approaches have been proposed.
 First attempts leverage \textit{local-to-global} techniques, where the collective behaviour emerges from local behaviour description, 
 usually inspired by nature~\cite{DBLP:journals/swarm/BrambillaFBD13}. 
%
Instead, a natural way to describe this kind of problems should be through a \textit{collective} view of the system.
 This is the idea used by the so-called \textit{global-to-local} techniques~\cite{DBLP:journals/jlap/ViroliBDACP19,DBLP:journals/scp/AlrahmanNL20, DBLP:conf/cbse/BuresGHKKP13}, 
 where the behaviours are designed \textit{top-down}, and then the local one is generated automatically from this ``system-wide'' specification. 
% 
Also ``meet-in-the-middle''~\cite{DBLP:journals/computer/PinciroliB16} approaches exist, where
 a part of the system is still expressed bottom-up (device-centric) whereas a part is expressed top-down (swarm-centric)

In this work, we focus on Aggregate Computing~\cite{DBLP:journals/computer/BealPV15}, a global-to-local approach where
 the self-organising collective behaviour is expressed in terms of manipulation of \textit{computational fields}.
The latter could be seen as the application of the mathematical concept of field: it consists of a 
 map from any devices to their computational value (i.e., the result of a computation).
The computational field manipulation is ruled by \textit{field calculus}, which define the constructs needed 
 to express spatio-temporal computations~\cite{DBLP:journals/jlap/ViroliBDACP19}.
%
The novelty of AC resides in its composable (functional-like) system-size independent program definition: 
 using basic constructs we identify the main patterns (\textit{building-blocks}) that could be reusable at application level.
By opportunistically crafting these building-blocks, it is possible to verify prominent properties such as eventual consistency %~\cite{DBLP:journals/taas/BealVPD17} 
 (i.e. the ensemble behaviour is independent of the underlying network details)
 and self-stabilisation %~\cite{DBLP:journals/tomacs/ViroliABDP18} 
 (i.e. the ability of a system to recover from arbitrary changes)~\cite{DBLP:journals/jlap/ViroliBDACP19}.

@@

\subsection{Field Calculus operators}
The main constructs that capture the essential aspects for programming self-organising systems with field calculus are:
\begin{itemize}
  \item \textit{Stateful field evolution} expression \mintinline[escapeinside=||]{scala}{|{\color{teal}rep}|(|$e_1$|) {|($x$) => $e_2$|}} describes a field evolving in time. 
  $e_1$ is the initial field value and the function $(x) => e_2$ define how the field changes at each execution.
  \item \textit{Neighbour interaction} expression \mintinline[escapeinside=||]{scala}{|\color{teal}nbr|{|$e$|}} a view
  of the field values in the surroundings of each device where neighbours are
  mapped to their evaluations of $e$. A series of \mintinline{scala}{*hood} operators leverage this map to produce relevant result. 
  For instance, \mintinline{scala}{minHood} returns the minimum value 
  within the neighbourhood.
  \item \textit{Domain partitioning} expression \mintinline[escapeinside=||]{scala}{|\color{teal}if|(|$e_0$|){|$e_1$|}{|$e_2$|}} splits the computational
  field into two non-communicating zones hosting isolated subcomputations:
  $e_1$ where $e_0$ is true, and $e_2$ where $e_0$ is false.
\end{itemize}
%
A full coverage of field calculus is beyond the scope of this paper. A more comprehensive presentation is given in~\cite{DBLP:journals/jlap/ViroliBDACP19}.

@@

In this case, when Reinforcement Learning is applied to several agents, we speak of Multi-Agent Reinforcement Learning (MARL)~\cite{DBLP:journals/aamas/Hernandez-LealK19}.
 AC has a typical cooperative and homogeneous learning setting: 
 each agent collaborates to achieve a common goal guided by the same reward function.
%
In the cooperative and homogeneous systems, several works have tried to apply the algorithms used in the single-agent context to the multi-agent one --- 
 the so-called Independent MARL~\cite{DBLP:journals/tsmc/BusoniuBS08}. 
%
However, several issues emerge, such as non-stationarity, multi-agent credit assignment, scalability and partial observability of the system. 
 Therefore, several heuristics have tried to solve these problems.
%
Among many, Hysteric Q-Learning~\cite{hysteretic-q} uses an optimistic approach in which more weight is given to positive actions than negative ones.
%
Another way to handle multi-agent learning is to use the so-called \textit{Centralised Training and Decentralised Execution} (CTDE) algorithms. 
 In this case, a central agent collects the experiences of the various agents and then distils a policy that each agent can execute locally in a typical actor-critic setting. 
%
During the learning phase, the problems of partial observability and non-stationarity are reduced, since the centralised learner has a global view. 
 However, the problem of scalability is typically not addressed.

@@

\section{Aggregate Computing with Reinforcement}\label{aggregate-and-rl}
\begin{figure}
  \centering
  \includegraphics[width=\linewidth]{img/algorithm-learning.pdf}
  \caption{This figure shows the main steps of RL in AC.
  Each agent starts with his copy of a Q-table. 
  Then, during the simulation, they refine them Q-tables changing Q values according to their experience (different colours mean different values). 
  In the end, a new $Q_{max}$ table will be created aggregating each agent table. 
  Finally, a new episode starts using this $Q_{max}$ as a starting Q-table.
  }
  \label{fig:aggregate-q-learning}
\end{figure}
%As first experiments, we leverage RL to improve building blocks. 
% 
In this section, we briefly describe how RL can be injected into the AC computation loop,
 and what RL we used with AC.

From a local point of view, AC program execution can be seen as:
$$
f : \textit{Context} \rightarrow \textit{Output}
$$
Where $\textit{Context}$ contains:
\begin{itemize}
  \item $\sigma(name) \rightarrow \textit{data}$: a function from sensor name to the perceived value;
  \item $\phi(\textit{ID}) \rightarrow \textit{Export}$: a function from id to neighbourhoods export data.
\end{itemize}
The computational field $\theta$ contains all output produced by the system.

As far as RL settings are concerned, we need to define the state function and the reward function.
%
The execution needs to be distributed, so the state will be a function from the local node context:
$$
K := \Omega(\textit{Context})
$$

Instead, reward function might be dependent on the learning technique, time, and problem. 
 Indeed, if we perform learning offline, it could be defined from the overall \textit{simulated} computational field:
$$
\textit{R}_{global} := \psi({\theta})
$$
If we use an online mode, it needs to be defined as a function to the local node context:
$$
\textit{R}_{local} := \psi(\textit{Context})
$$
Finally, actions influence the current aggregate program output, changing the program evaluation as:
$$
f_{rl} : \textit{Context} \rightarrow \lambda(\textit{Output}, \textit{Action})
$$
As remarked before, the policy $\pi$ needs to be distributed, so that each agent can choose the action that will be performed:
$$
\textit{Action} \leftarrow \pi(K)
$$

As a first step towards a complete integration between AC and RL, we have decided to adopt an Independent RL learning technique, 
 by which each agent try to improve their own current local policy (shown in \Cref{fig:aggregate-q-learning}). 
% 
We made this choice inspired by other work in CSAS~\cite{csas-and-marl} and swarm systems~\cite{nguyen2018swarm}.

Our approach is partially inspired by Swarm Q-Learning~\cite{nguyen2018swarm} and leverage Hysteretic Q-Learning~\cite{hysteretic-q} to
 handle the non-stationarity in the environment.
%
In particular, each agent updates a local Q-table ($Q_i(s, a)$, where $i$ is the agent id) following the standard hysteretic update:
$$
\delta \leftarrow r + \gamma * max_a Q_i(s', a) - Q_i(s_t, a_t)
$$
$$
Q(s_t, a_t) =  \begin{cases} 
  Q(s_t, a_t) + \alpha * \delta & \mbox{if } delta >= 0 \\ 
  Q(s_t, a_t) + \beta * \delta & \mbox{otherwise }
\end{cases}
$$

At the end of each episode, each agent collects its local experience, which can vary in different zones. 
 So, we combine the $N$ (the number of nodes) Q-table selecting the maximum value for each state-action pair since we assume an optimistic homogeneous behaviour:
$$
Q_{max}(s, a) = max_i(Q_i(s, a))
$$
Then, in the next episode, each agent starts with the $Q_{max}$ and try to refine it.
\section{Evaluation}\label{evaluation}
\begin{figure*}[h]
  \centering
  \begin{subfigure}[b]{0.32\textwidth}
      \centering
      \includegraphics[width=\textwidth]{img/hop-count-1.png}
      \caption{Stable situation: Two source nodes exists}
  \end{subfigure}
  \hfill
  \begin{subfigure}[b]{0.32\textwidth}
      \centering
      \includegraphics[width=\textwidth]{img/hop-count-2.png}
      \caption{Unstable situation: One source disappears}
  \end{subfigure}
  \hfill
  \begin{subfigure}[b]{0.32\textwidth}
      \centering
      \includegraphics[width=\textwidth]{img/hop-count-3.png}
      \caption{Stable state reached}
  \end{subfigure}
  \caption{Graphical result of hop count evaluation. The node with a red cross is the target zone. The label 
  associated with each node is the result of the aggregate program evaluated in that zone.}
  \label{fig:hop-count}
\end{figure*}
In this section, we outline the results of the application of RL for one simple
 aggregate program: the hop count. In this case, the computational field that will eventually be produced consists of the distance in terms of the minimum number of nodes needed to reach source nodes (\Cref{fig:hop-count}).

@@

Where \textbf{\texttt{k}} and \textbf{\texttt{w}} are two parameters of this algorithm.
%
The policy is learnt locally, during simulation, using our version of Q-Learning algorithm. 
 The reward function consists in: 
\[
\systeme*{0 : & \mbox{\textit{hop count output is right}}, -1 : & \mbox{\textit{otherwise}}}
\]
The idea is to maintain the nodes in a stable situation or to
 speed up the time needed to reach a stable one.
%
\Cref{fig:aggregate-learning-loop} shows the learning loop in details.
\begin{figure}
  \centering
  \includegraphics[width=\linewidth]{img/aggregate-program-evaluation.pdf}
  \caption{Each agent, starting from their context, 
  periodically performs the standard hop count. 
  Using the output produced and the history of state and output, 
  each agent computes its current state. 
  Then, an agent evaluates the action to speed up the hop count convergence following a policy. 
  This new refined output is then used to evaluate the reward and to improve the Q-table (dotted line path)}
  \label{fig:aggregate-learning-loop}
\end{figure}

\subsection{Simulation Setup}
We verify our learning algorithm through simulation\footnote{repository at \url{https://github.com/cric96/alpaca-2022-ac-rl}} leveraging Alchemist and ScaFi.
 To check if the aggregate eventually speeds up the hop count convergence time, we create a situation where the rising problem emerges (\Cref{fig:hop-count}):
 firstly, two source nodes exist in the network, ($A$ and $B$).
 At the time $t$, $B$ disappears. So the nodes near $B$ start a slow process
 to eventually reach the distance towards $A$.
We use 80 ($N$) nodes displaced into an irregular grid with a width of 100 meters and a height of 20 meters. 
 Each node has a neighbourhood range equal to 8 meters --- on average, each node has 7 neighbours. 
 The time $t$ is fixed to 20.
 Each simulation episode lasts 60 time-unit. Finally, we set $k = 3$ and $w = 2$.
% 
We train the aggregate for 500 episodes.
 The Q-learning parameters are:
$$ 
\alpha = 0.5,
\beta = 0.05,
\gamma = 0.9
$$
Moreover, we use a $\epsilon$ decay policy for the $\epsilon$-greedy policy as:
\begin{equation*}
  \begin{array}{l}
    \epsilon_t = \epsilon_{start} * e^{-t/\theta}\\ 
    \epsilon_{start }= 0.99, \theta=4000
  \end{array}
\end{equation*}
In doing so, the agent will explore the state-action space as much as possible.

@@

Then, we test our learnt algorithm in different environments (\Cref{fig:simulation} b, c, d). To do that, we share the learnt Q-table to the 
 entire aggregate and we test the program changing the system size (80, 360 and 760 nodes).
%
For each population size, we run 50 episodes and then we average the error at each time step. 
The aggregate speed-up convergence is observed for each scenario taken into consideration.

@@

\section{Discussion and Concluding Remarks}\label{conclusion}
In this article, we have discussed the combination of AC with Machine Learning techniques in order to improve adaptivity
 and effectiveness of the AC framework.
In particular, we explore the combination of AC with Hysteretic Q-Learning.
The application of RL to this kind of setting is a largely unexplored %-- even if some works exists~\cite{DBLP:conf/icml/YangLLZZW18} -- 
 research area. There are a variety of challenges, such as non-stationarity,  partial environment observability, and non-bounded system size.
In a case study, we have empirically showed the effectiveness of RL by showing a speed up of the convergence time of
 aggregate program we took into consideration, namely the hop count.
Furthermore, we have shown that the learnt program can easily scale up to system size.

To the best of our knowledge, this is the first effort towards a ``hybrid'' collective programming approach, where the high-level program specification is 
 done by a designer and the low-level implementation to building blocks based on machine learning.
%
As such, this work has shed light on several issues deserving future investigation.

Aggregate systems are partially observable from the perspective of each node. 
 Local observations may not be statistically significant in summarising the state of the entire aggregate. 
 In the future, we could address this problem through recurrent neural networks and a CTDE learning methodology.
 Learning algorithms were implemented at building block level. 
 Another possible direction is to use them at the framework level, e.g., to select which messages to send to the neighbourhood or not. 
 This has already been addressed in the literature under the name of \textit{learn to communicate}~\cite{DBLP:journals/corr/abs-1908-03963}. 
A novel way to handle this problem is to use Graph Neural Networks~\cite{DBLP:journals/corr/abs-1812-08434} to learn the representation of a distributed system and to select messages based on that representation. 
 Thus, we should consider this methodology when we will extend RL at the AC framework level.

In our experiments, we have applied a variation of the independent learning approach. 
 Agents first learn the best policy locally, and then at the end of the episode, a learner merges the Q-tables. 
 However, this does not reduce the problems of scalability and non-stationary, since a central agent must still collect all the Q-tables of the aggregate. 
 In particular, to handle the scalability issues, we will have to consider decentralised approaches based only on neighbour data (or local leaders).
